---
title: "Exploratory data analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EDA}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}

library(dplyr)
library(ggplot2)
library(knitr)
library(skimr)
library(stringr)
library(tidyr)

library(brazildatamortality)


```


```{r check_variables, include=FALSE}

# Check that data before and after 1096 have the same variables and categories.

data_9  <- brazildatamortality::data_icd_9
data_10 <- brazildatamortality::data_icd_10

colnames(data_10) %>% sort()
colnames(data_9)  %>% sort()

# OK
data_10 %>% sample_n(6) %>% pull(birth_date) %>% unique() %>% sort()
data_9  %>% sample_n(6) %>% pull(birth_date) %>% unique() %>% sort()

# NOTE: They don't math too much because data before 1996 is deficient. 
data_10 %>% pull(cause) %>% unique() %>% sort()
data_9  %>% pull(cause) %>% unique() %>% sort()

# NOTE: 6 versus 7 digit codes. 
data_10 %>% sample_n(6) %>% pull(residence_city) %>% unique() %>% sort()
data_9  %>% sample_n(6) %>% pull(residence_city) %>% unique() %>% sort()

# NOTE: 6 versus 7 digit codes. 
data_10 %>% sample_n(6) %>% pull(death_city) %>% unique() %>% sort()
data_9  %>% sample_n(6) %>% pull(death_city) %>% unique() %>% sort()

# OK
data_10 %>% sample_n(6) %>% pull(death_date) %>% unique() %>% sort()
data_9  %>% sample_n(6) %>% pull(death_date) %>% unique() %>% sort()

# TODO: Recode jobs.
data_10 %>% sample_n(6) %>% pull(job) %>% unique() %>% sort()
data_9  %>% sample_n(6) %>% pull(job) %>% unique() %>% sort()

# TODO: Issue with Bachelor incomplete. 
data_10 %>% pull(education) %>% unique() %>% sort()
data_9  %>% pull(education) %>% unique() %>% sort()

# TODO: What happened to hospital? (sid 10)
data_10 %>% sample_n(6) %>% pull(locus) %>% unique() %>% sort()
data_9  %>% sample_n(6) %>% pull(locus) %>% unique() %>% sort()

# TODO: No information, other, ignored
data_10 %>% pull(marital) %>% unique() %>% sort()
data_9  %>% pull(marital) %>% unique() %>% sort()

# OK
data_10 %>% pull(sex) %>% unique() %>% sort()
data_9  %>% pull(sex) %>% unique() %>% sort()

# TODO: Add to the package the list of codes of towns from IBGE (SHP).

```



```{r merge_both_data, include=FALSE}

data_10 <- data_10 %>% 
  dplyr::select(birth_date, cause, death_city, death_date, education, 
                job, locus, marital, residence_city, sex)
data_9  <- data_9  %>% 
  dplyr::select(birth_date, cause, death_city, death_date, education, 
                job, locus, marital, residence_city, sex)

data_tb <- data_9 %>%
  dplyr::bind_rows(data_10) %>%
  dplyr::mutate(death_year = lubridate::year(death_date),
                computed_age = difftime(death_date, birth_date, 
                                        units = "days"),
                death_month = lubridate::month(death_date)) %>%
  tidyr::drop_na(cause, death_date)

outlier_data <- data_tb %>%
  dplyr::filter(death_date > as.Date("2011-01-10"),
                death_date < as.Date("2011-01-13"))

regular_data <- data_tb %>%
  dplyr::filter(death_date < as.Date("2011-01-11") | 
                death_date > as.Date("2011-01-12"))

```



## Abstract.

## Introduction.

## Materials and methods.

## Results.

```{r eda, echo=FALSE, fig.width = 7, fig.height = 5}

#library(GGally)

# data_tb %>%
#   dplyr::filter(death_year > 2005) %>%
#   dplyr::mutate(death_month = stringr::str_pad(death_month, pad = "0", width = 2)) %>%
#   dplyr::mutate(death_year = as.factor(death_year),
#                 death_month = as.factor(death_month)) %>%
#   dplyr::select(cause, education, locus, marital, 
#                 sex, death_year, death_month) %>%
#   GGally::ggpairs(ggplot2::aes(color = cause))

# data_tb %>%
#   dplyr::mutate(death_month = stringr::str_pad(death_month, pad = "0", 
#                                                width = 2)) %>%
#   dplyr::mutate(death_year = as.factor(death_year),
#                 death_month = as.factor(death_month)) %>%
#   dplyr::select(cause, education, locus, marital, 
#                 sex, death_year, death_month) %>%
#   tidyr::pivot_longer(education:death_month) %>%
#   ggplot2::ggplot(ggplot2::aes(y = value, fill = cause)) +
#   ggplot2::geom_bar(position = "fill") +
#   ggplot2::facet_wrap(vars(name), scales = "free") +
#   ggplot2::labs(x = NULL, y = NULL, fill = NULL)


```


```{r plot_cumsum, echo=FALSE, fig.width = 7, fig.height = 5}

regular_data %>%
  dplyr::group_by(death_year) %>%
  dplyr::summarize(fatalities = dplyr::n(), 
                   .groups = "drop") %>%
  dplyr::mutate(cumsum = cumsum(fatalities)) %>%
  dplyr::arrange(death_year) %>%
  ggplot2::ggplot(ggplot2::aes(x = death_year,
                               y = cumsum)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::xlab("Year") +
  ggplot2::ylab("Accumulated number of fatalities") +
  ggplot2::ggtitle("Cumulative number of deaths by forces of nature")


```

```{r plot_causes_cumsum, echo=FALSE, fig.width = 7, fig.height = 5}

regular_data %>%
  dplyr::group_by(cause, death_year) %>%
  dplyr::summarize(fatalities = dplyr::n()) %>%
  dplyr::mutate(cumsum = cumsum(fatalities)) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(cause, death_year) %>%
  ggplot2::ggplot(ggplot2::aes(x = death_year,
                               y = cumsum,
                               color = cause,
                               group = cause)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::xlab("Year") +
  ggplot2::ylab("Cumulative number of fatalities") +
  ggplot2::ggtitle("Cumulative sum of deaths by cause")

```

```{r plot_total, echo=FALSE, fig.width = 7, fig.height = 5}

regular_data %>%
  dplyr::group_by(death_year) %>%
  dplyr::summarize(fatalities = dplyr::n(), 
                   .groups = "drop") %>%
  dplyr::mutate(moving_average = RcppRoll::roll_mean(fatalities, 
                                                     n = 5, 
                                                     align = "right", 
                                                     fill = NA,
                                                     na.rm = TRUE))  %>%
  dplyr::arrange(death_year) %>%
  ggplot2::ggplot(ggplot2::aes(x = death_year,
                               y = fatalities)) +
  ggplot2::geom_point() +
  ggplot2::geom_line(ggplot2::aes(x = death_year,
                                  y = moving_average),
                                  color = "blue") +
  ggplot2::xlab("Year") +
  ggplot2::ylab("Number of fatalities") +
  ggplot2::ggtitle("Total number of deaths by forces of nature")

```

```{r plot_causes_sex_cumsum, echo=FALSE, fig.width = 7, fig.height = 5}

regular_data %>%
  dplyr::filter(sex %in% c("Male", "Female")) %>%
  dplyr::group_by(sex, cause, death_year) %>%
  dplyr::arrange(sex, cause, death_year) %>%
  dplyr::summarize(fatalities = dplyr::n()) %>%
  dplyr::filter(cause %in% c("Cataclismyc", 
                             "Cataclismyc and Floods",
                             "Cold", 
                             "Lightning", 
                             "Landslide")) %>%
  dplyr::mutate(cumsum = cumsum(fatalities)) %>%
  dplyr::ungroup() %>%
  ggplot2::ggplot(ggplot2::aes(x = death_year,
                               y = cumsum,
                               color = cause,
                               group = cause)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::facet_wrap(~sex) +
  ggplot2::xlab("Year") +
  ggplot2::ylab("Cumulative number of fatalities by sex") +
  ggplot2::ggtitle("Cumulative sum of deaths by cause and sex")

regular_data %>%
  dplyr::filter(sex %in% c("Male", "Female")) %>%
  dplyr::group_by(sex, cause, death_year) %>%
  dplyr::arrange(sex, cause, death_year) %>%
  dplyr::summarize(fatalities = dplyr::n()) %>%
  dplyr::filter(!(cause %in% c("Cataclismyc", 
                             "Cataclismyc and Floods",
                             "Cold", 
                             "Lightning", 
                             "Landslide"))) %>%
  dplyr::mutate(cumsum = cumsum(fatalities)) %>%
  dplyr::ungroup() %>%
  ggplot2::ggplot(ggplot2::aes(x = death_year,
                               y = cumsum,
                               color = cause,
                               group = cause)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::facet_wrap(~sex) +
  ggplot2::xlab("Year") +
  ggplot2::ylab("Cumulative number of fatalities by sex") +
  ggplot2::ggtitle("Cumulative sum of deaths by cause and sex")

```

```{r plot_causes, echo=FALSE, fig.width = 7, fig.height = 5}

# High impact
data_tb %>%
  dplyr::group_by(cause, death_year) %>%
  dplyr::summarize(fatalities = dplyr::n(), 
                   .groups = "drop") %>%
  dplyr::arrange(cause, death_year) %>%
  dplyr::filter(cause %in% c("Cataclismyc",
                             "Cataclismyc and Floods", 
                             "Flood",
                             "Landslide",
                             "Lightning")) %>%             
  dplyr::filter(!is.na(cause), !is.na(death_year)) %>%
  ggplot2::ggplot(ggplot2::aes(x = death_year,
                               y = fatalities,
                               color = cause,
                               group = cause)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::xlab("Year") +
  ggplot2::ylab("Number of fatalities") +
  ggplot2::ggtitle("Death by cause") + 
  ggplot2::scale_color_brewer(palette = "Paired")

# Low impact
data_tb %>%
  dplyr::group_by(cause, death_year) %>%
  dplyr::summarize(fatalities = dplyr::n(), 
                   .groups = "drop") %>%
  dplyr::arrange(cause, death_year) %>%
  dplyr::filter(!(cause %in% c("Cataclismyc",
                             "Cataclismyc and Floods", 
                             "Flood",
                             "Landslide",
                             "Lightning"))) %>%             
  dplyr::filter(!is.na(cause), !is.na(death_year)) %>%
  ggplot2::ggplot(ggplot2::aes(x = death_year,
                               y = fatalities,
                               color = cause,
                               group = cause)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::xlab("Year") +
  ggplot2::ylab("Number of fatalities") +
  ggplot2::ggtitle("Death by cause") + 
  ggplot2::scale_color_brewer(palette = "Paired")

```



```{r plot_causes_esc_cumsum, echo=FALSE, fig.width = 7, fig.height = 5}

# TODO: What do we do with the GAP from 1995 to 2012?
# The variables of CID9 and CID10 are not the same
regular_data %>%
  dplyr::mutate(education = as.factor(education),
                cause     = as.factor(cause)) %>%
  tidyr::drop_na(education) %>%
  dplyr::group_by(education, cause, death_year) %>%
  dplyr::arrange(education, cause, death_year) %>%
  dplyr::summarize(fatalities = dplyr::n()) %>%
  dplyr::mutate(cumsum = cumsum(fatalities)) %>%
  dplyr::ungroup() %>%
  ggplot2::ggplot(ggplot2::aes(x = death_year,
                               y = cumsum,
                               color = cause,
                               group = cause)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  #ggplot2::scale_color_brewer(palette = "Spectral") + 
  #ggplot2::scale_colour_viridis_d(option = "plasma") +
  ggplot2::facet_wrap(~education) +
  ggplot2::xlab("Year") +
  ggplot2::ylab("Cumulative number of fatalities by education") +
  ggplot2::ggtitle("Cumulative sum of deaths by cause and education")

```


```{r plot_by_cause, echo=FALSE, fig.width = 7, fig.height = 5}

regular_data %>%
  dplyr::group_by(death_year, cause) %>%
  dplyr::summarize(fatalities = log(dplyr::n()),
                   .groups = "drop") %>%
  # NOTE: Util function for adding the roll means as a new column.
  (function(x) {
    my_data <- x %>%
      dplyr::arrange(death_year, cause) %>%
      tidyr::pivot_wider(id_cols = death_year,
                         names_from = cause,
                         values_from = fatalities)
    my_roll <- RcppRoll::roll_mean(as.matrix(dplyr::select(my_data, 
                                                           -death_year)),
                                   n = 5,
                                   align = "right",
                                   fill = NA,
                                   na.rm = TRUE) %>%
      tibble::as_tibble() %>%
      magrittr::set_names(paste(names(.), "roll_mean", sep = "_")) %>%
      dplyr::mutate(death_year = my_data$death_year) %>%
      tidyr::pivot_longer(!death_year) %>%
      dplyr::rename(moving_average = value)
   my_data %>%
     tidyr::pivot_longer(!death_year,
                         names_to = "cause",
                         values_to = "fatalities") %>%
     dplyr::bind_cols(dplyr::select(my_roll, moving_average)) %>%
     return()
  }) %>%
  dplyr::arrange(death_year) %>%
  ggplot2::ggplot(ggplot2::aes(x = death_year,
                               y = fatalities)) +
  ggplot2::geom_point() +
  ggplot2::geom_line(ggplot2::aes(x = death_year,
                                  y = moving_average),
                                  color = "blue") +
  ggplot2::facet_wrap(~cause,
                      ncol = 3) +
  ggplot2::xlab("Year") +
  ggplot2::theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggplot2::ylab("Logarithm of the number of fatalities") +
  ggplot2::ggtitle("Number of deaths by forces of nature by cause")

```

```{r cause_state, echo=FALSE, fig.width = 10, fig.height = 7}

library(geobr)
library(sf)

# Download the data.
state_sf <- geobr::read_state(year = 2010, 
                              simplified = TRUE, 
                              showProgress = FALSE)
state_tb <- state_sf %>%
  sf::st_set_geometry(NULL) %>%
  tibble::as_tibble()

town_sf <- geobr::read_municipality(year = 2010, 
                                    simplified = TRUE,
                                    showProgress = FALSE)
town_tb <- town_sf %>%
  sf::st_set_geometry(NULL) %>%
  tibble::as_tibble() %>%
  dplyr::mutate(code_muni = as.character(code_muni)) %>%
  dplyr::left_join(state_tb, by = "abbrev_state")

data_tb %>%
  dplyr::left_join(town_tb, by = c("death_city" = "code_muni")) %>%
  dplyr::group_by(death_year, cause, abbrev_state) %>%
  dplyr::summarize(fatalities = dplyr::n(), 
                   .groups = "drop") %>%
  tidyr::drop_na() %>%
  ggplot2::ggplot(ggplot2::aes(x = death_year,
                               y = fatalities, 
                               color = cause, 
                               group = cause)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::facet_wrap(vars(abbrev_state)) +
  ggplot2::xlab("Year") +
  ggplot2::ylab("Number of fatalities") +
  ggplot2::ggtitle("Deaths by cause by state")

data_tb %>%
  dplyr::left_join(town_tb, by = c("death_city" = "code_muni")) %>%
  dplyr::group_by(death_year, cause, name_region) %>%
  dplyr::summarize(fatalities = dplyr::n(), 
                   .groups = "drop") %>%
  tidyr::drop_na() %>%
  ggplot2::ggplot(ggplot2::aes(x = death_year,
                               y = fatalities, 
                               color = cause, 
                               group = cause)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::facet_wrap(vars(name_region)) +
  ggplot2::xlab("Year") +
  ggplot2::ylab("Number of fatalities") +
  ggplot2::ggtitle("Deaths by cause by region")

```


TODO: make figure including
+ data_10 cataclismyc, flood, and landslide
+ data_9 cataclismyc & floods, and Earth surface movement.
+ Add figure of low impact death causes by season (x = month, 
y = number of fatalities, color = cause).
+ design sex-analysis figure for  outlier_data NOT using time series.
+ maps by region? by state? which year?

## Conclusions.

## References.
